Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-05-27 Manuscript/manuscript.tex
More discussion edits.

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
Thus, when high quality habitat is preferentially selected by migrants (i.e. $\xi_{em} > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

When pathogen spread is weak, high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.  The shift in this trade-off can arise from conditions that change either side of that balance.  High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.

The trade-off can also switch in cases that disrupt the positive benefits of high quality habitat for metapopulation stability, i.e. pathogens with high disease-induced mortality or when dispersal is limited by metapopulation structure.  In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.

As pathogen longevity increases and environmental transmission becomes more important relative to direct transmission (i.e. the top right of the "u") the trade-off shifts back and the addition of high quality patches is again a net benefit to overall occupancy.  In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.  Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.

>>>>

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
Thus, when high quality habitat is preferentially selected by migrants (i.e. $\xi_{em} > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Fig~\ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Thus, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).  
Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.  
High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.

In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.  More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.  Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.
